---
title: "Data Connect Library - Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Connect Library - Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Prerequisite: Install R & RStudio if you have not installed before.**

#### Install devtools, if it was not already installed before.

```{r eval=FALSE}

install.packages("devtools")

```

#### Install reticulate, if it was not already installed before.

```{r eval=FALSE}

install.packages("reticulate")

```

#### Install dataconnect library from GitHub

```{r eval=FALSE}

try({remove.packages("dataconnect")}, silent=TRUE)

devtools::install_github(
  repo = "mdsol/dataconnect-library-r", 
  ref = "v1.1.0",
  build_vignettes = TRUE, 
  upgrade = FALSE)

```

#### Setup virtual environment, only if it was not created before. It is not required to run this step when updated package is installed.

```{r eval=FALSE}

library(dataconnect)
install_miniforge()

```

### Windows users: If you are using Windows, please run the following command in R to set up the Python environment for the dataconnect library. This step is not required for Linux or MacOS users.

```{r eval=FALSE}
Sys.setenv(RETICULATE_PYTHON = "C:/Users/YOUR_USERNAME/miniforge3/envs/dataconnect-library-r/python.exe")

library(reticulate)
use_python("C:/Users/YOUR_USERNAME/miniforge3/envs/dataconnect-library-r/python.exe", required = TRUE)
```


***For usage, please refer this [guide](rLibrary_usage.html).***

