---
title: "Data Connect Library - Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Connect Library - Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
***Version Note:***
This vignette reflects the *latest version* of the `dataconnect` R library. 
If you are using an older version, the functions or parameters described here may differ from your installed version.

***View version-matched documentation:***
On GitHub, use the branch/tag selector to switch to your version's tag (e.g., `v1.0.1`).

### Check your version

```{r eval=FALSE}

packageVersion("dataconnect")
```
***To set up Data Connect Library, please follow these [instructions](rLibrary_setup.html).***

### Preparation

Please restart R session in RStudio before proceeding, if you have just installed the dataconnect R library and set up the conda environment.

```{r eval=FALSE}
# We recommend storing userToken in a separate file in your working directory.
# This is the user authentication token generated from the Developer Center in iMedidata Platform.

userToken = "usertoken"
```

```{r eval=FALSE}
# Load prerequisite libraries
library(tidyverse)
```

### Initialize Data Connect connection

```{r eval=FALSE}
library(dataconnect)
use_miniforge_env()
  
dc <-init(
  token = userToken)

```

### Get all available studies

Get all available studies or search for a study by name.

```{r eval=FALSE}

tryCatch({
  result <- dc$studies(search_study_name = "", page = 1, page_size = 10)

  cat(sprintf("\nPage: %s, Page size: %s, Total pages: %s\n",
              result$pagination$page,
              result$pagination$page_size,
              result$pagination$total_pages))

  cat(sprintf("\n--- Showing %d of %d Studies ---\n\n", length(result$studies),
              result$total_records))

  for (study in result$studies) {
    envs <- sapply(study$environments, function(e) e$name)
    envs_list <- paste(envs[nzchar(envs)], collapse = ", ")

    cat(sprintf("• Study: %s\n", study$name))
    if (nzchar(envs_list)) {
      cat(sprintf("  Envs:  %s\n\n", envs_list))
    } else {
      cat("  Envs:  (none)\n\n")
    }
  }

}, dataconnect_error = function(e) {

  # Catch dataconnect specific errors.
  # Instead of printing the entire error object, we can extract and print specific details such as the error message and code.
  print(e)

}, error = function(e) {

  # Catch any other errors
  cat("Error:", e$message, "\n")

})
```

### Work with datasets with `dataconnect` library

#### Search datasets in a study environment

Retrieve the study and study environment information from "Developer info" in iMedidata Platform.

```{r eval=FALSE}

PWBStudy = list(
  PWBStudyId = "studyid",
  PWBPRODEnvID = "studyenvironmentid"
)

ExciterStudy = list (
  ExciterStudyId = "studyid",
  ExciterPRODEnvID = "studyenvironmentid"
)
```

Set search keyword.

```{r eval=FALSE}
searchName = "datasetname"
```

Search for a dataset in a specific study environment.

```{r eval=FALSE}

tryCatch({

  datasets <- dc$datasets(
    study_environment_uuid = ExciterStudy$ExciterPRODEnvID,
    search_dataset_name = searchName,
    page = 1,
    page_size = 5
  )

  print(paste("Total datasets available across all pages:", datasets$total_records))

  print(sprintf("Page: %s, Page size: %s, Total pages: %s",
                datasets$pagination$page,
                datasets$pagination$page_size,
                datasets$pagination$total_pages))

  for (dataset in datasets$datasets) {
    print(to_frame(dataset))
  }

}, dataconnect_error = function(e) {

  # Catch dataconnect specific errors.
  # Instead of printing the entire error object, we can extract and print specific details such as the error message and code.
  print(e)

}, error = function(e) {

  # Catch any other errors
  cat(e$message, "\n")
})

```

#### Identify different versions of a specific dataset

Prepare a list of datasets that are needed.

```{r eval=FALSE}
# datasetid can be retrieved from search datasets()
datasetid = list(
  PWBDerID = "datasetid",
  PWBImportID = "datasetid",
  PWBRaveID = "datasetid",
  PWBCustomID = "datasetid",
  ExciterDerID = "datasetid",
  ExciterImportID = "datasetid",
  ExciterRaveID = "datasetid",
  ExciterCustomID = "datasetid"
)
```

Retrieve other dataset versions if they are available.

```{r eval=FALSE}

tryCatch({

  dataset_versions <- dc$dataset_versions(
    dataset_uuid = datasetid$ExciterCustomID
  )

  for (dataset_version in dataset_versions) {
    print(to_frame(dataset_version))
  }

}, dataconnect_error = function(e) {

  # Catch dataconnect specific errors.
  # Instead of printing the entire error object, we can extract and print specific details such as the error message and code.
  print(e)

}, error = function(e) {

  # Catch any other errors
  cat(e$message, "\n")
})
```

#### Fetch records for specific datasets

```{r eval=FALSE}
tryCatch({
  RAVEdata <- dc$fetch_data(
                  dataset_uuid = datasetid$ExciterRaveID
  )
  # Fetching first 3 rows of data
  RAVEdata$frame %>% head(3)

}, dataconnect_error = function(e) {
  print(e)
}, error = function(e) {
  cat(e$message, "\n")
})

```

```{r eval=FALSE}
tryCatch({
  Derdata <- dc$fetch_data(
                  dataset_uuid = datasetid$PWBDerID
  )
  # Fetching the first 6 rows of data
  Derdata$frame %>% head()

}, dataconnect_error = function(e) {
  print(e)
}, error = function(e) {
  cat(e$message, "\n")
})
```

```{r eval=FALSE}
tryCatch({
  ImportData <- dc$fetch_data(
                  dataset_uuid = datasetid$PWBImportID
  )
  
  # Fetching the first 3 rows of data
  ImportData$frame %>% head(3)
}, dataconnect_error = function(e) {
  print(e)
}, error = function(e){
  cat(e$message, "\n")
})
```

```{r eval=FALSE}
tryCatch({
  CustomData <- dc$fetch_data(
                  dataset_uuid = datasetid$ExciterCustomID
  )
  
  # Fetching the first 3 rows of data
  CustomData$frame %>% head(3)

}, dataconnect_error = function(e) {
  print(e)
}, error = function(e)
{
  cat(e$message, "\n")
})
```

```{r eval=FALSE}
# MATERIALIZE LOCALLY: bring the data into memory as a data.frame
# collect() should return the entirety of the dataset 

Derdf = Derdata$frame %>% collect()
Importdf = ImportData$frame %>% collect()
Customdf = CustomData$frame %>% collect()
```

```{r eval=FALSE}
# Consider fetching the first few rows of a dataset by using %>% head() if the dataset is extremely large

RAVEdf = RAVEdata$frame %>% head()
```

### Data Transformation Example Using Dplyr

#### Data transformation using native dplyr capability in R

Materialize the data into the client side memory as a R dataframe.

If you have limited memory allocation, we recommend working on a limited set of records to reduce development time, and then remove record limit during recurring execution.

If you receive an error indicating that you are running out of memory, please reach out to your R IDE provider to increase the memory allocation.

#### Set up: Using dplyr to perform data transformation on materialized data frame

These functions are not provided by the `dataconnect` library. Below are just examples of how common R libraries can be used with the `dataconnect` library. For details on how these libraries can be used, please consult the respective library documentation.

```{r dplyr-basic, eval=FALSE}
# Pivot Derdf as pivotDF
pivotDF <- Derdf %>%
  rename(
    HCT_result = HCT,
    plat_result = PLAT,
    wbc_result = WBC,
    HCT_unit = HCT_UNIT,
    plat_unit = PLAT_UNIT,
    wbc_unit = WBC_UNIT
  ) %>%
  pivot_longer(
    cols = -c(patient_id, site_id, LBTIM),
    names_to = c("lab_test", ".value"),
    names_sep = "_"
  ) %>%
  rename(
    lab_result = result,
    lab_unit = unit
  )

# Union the datasets together (pivotDF, Importdf) as unionDF
pivotdf_renamed <- pivotDF %>%
  rename(
    subjid = patient_id,
    siteid = site_id,
    lbtest = lab_test,
    lbresn = lab_result,
    lbresu = lab_unit
  )
unionDF <- bind_rows(Importdf, pivotdf_renamed)

# Join the datasets together (unionDF, Customdf) as publishDF
x = Importdf
y = unionDF
publishDF <- full_join(
  x,
  y %>% select(subjid, 
               siteid, 
               visitnum, 
               -any_of(
                 setdiff(
                   names(x),
                   c("subjid", "siteid", "visitnum")
                   )
                 )
               ),
  by = c("subjid" = "subjid", "siteid" = "siteid", "visitnum" = "visitnum")
)
```

### Publishing

Set up your project token where the resulting dataset (i.e., publishDF) will be published to. Please make sure it is published to the correct study environment and project. It is recommended to use dry_publish function to validate the result first before publishing the resulting dataset back to Medidata Data Connect.

```{r eval=FALSE}
# The project token is generated in the Data Connect Transformation Custom Code Project

myProjectToken = "projecttoken"
```

### Set Datetime formats for the columns with date or timestamp data types
```{r}
# Set datetime formats for the columns with date or timestamp data types. The format should follow Java DateTimeFormatter patterns.
# For example, if start_date column is in "MM/dd/yy" format, end_date column is in "yyyy-MM-dd" format, and update_date column is in "yyyy-MM-dd HH:mm:ss" format, the datetime_formats list should be set as below:
datetime_formats_list <- list("start_date" = "MM/dd/yy", "end_date" = "yyyy-MM-dd", "update_date" = "yyyy-MM-dd HH:mm:ss")
```

#### Validate the dataset to be published

```{r eval=FALSE}
# Validating the published dataset. Comment out when set on a recurring schedule.
tryCatch({
  dry_publish_result <- dc$dry_publish(
    project_token = myProjectToken,
    dataset_name = "projectname",
    key_columns = list ("field1","field2", "field3"),
    source_datasets = list(datasetid$PWBDerID, datasetid$PWBImportID, datasetid$PWBRaveID),
    data = publishDF,
    datetime_formats = datetime_formats_list
  )
  
  print(dry_publish_result)
  
}, dataconnect_error = function(e){
  
  # Catch dataconnect specific errors. 
  # Instead of printing the entire error object, we can extract and print specific details such as the error message and code.
  print(e)

}, error = function(e) {

  # Catch any other errors
  cat(e$message, "\n")
})

```

#### Publish

```{r eval=FALSE}
tryCatch({
  publish_result <- dc$publish(
    project_token = myProjectToken,
    dataset_name = "projectname",
    key_columns = list ("field1","field2", "field3"),
    source_datasets = list(datasetid$PWBDerID, datasetid$PWBImportID, datasetid$PWBRaveID),
    data = publishDF,
    datetime_formats = datetime_formats_list
  )

  print(publish_result)
  
}, dataconnect_error = function(e){
  
  # Catch dataconnect specific errors. 
  # Instead of printing the entire error object, we can extract and print specific details such as the error message and code.
  print(e)

}, error = function(e) {

  # Catch any other errors
  cat(e$message, "\n")
})
```
